---
title: "Enhance Your Data Science Workflow with the box R Package"
author: "Alier Reng"
date: "04/16/2023"
format: html
editor: visual
---

## Tutorial on Utilizing the Box R Package Effectively

This tutorial demonstrates how to incorporate the box R package into your data science workflow effectively ðŸ–¥ï¸.

## Outline of this Tutorial

1.  What is box?

> \'box\' allows organising R code in a more modular way, via two mechanisms:
>
> 1.  It enables *writing modular code* by treating files and folders of R code as independent (potentially nested) modules, without requiring the user to wrap reusable code into packages.
>
> 2.  It provides a new syntax to import reusable code (both from packages and from modules) which is more powerful and less error-prone than **`library`** or **`require`**, by limiting the number of names that are made available.
>     ([The Box R Package](https://morioh.com/p/9a313401f051))

2.  Loading the Package

    option + command + i

```{r}
#| warning: false
#| message: false

# Libraries
# library(tidyverse)
box::use(
  dplyr[...], 
  purrr[map_df], 
  tidyr[...], 
  stringr[str_remove_all],
  readxl[read_excel, read_xlsx]
)

```

2.  Importing the Dataset

```{r}
# Importing the dataset
file_path <- fs::dir_ls("../00-input/excel_files/")

census <- file_path |>
  
  # map_df(read_xlsx)
  # map_df(file_path, ~read_excel(.x))
  map_df(\(path) read_xlsx(path)) |> 
  
  # Select columns of interest
  janitor::clean_names() |> 
  select(
    former_region, 
    state = region_name,
    gender = variable_name,
    age_category = age_name,
    population = x2008
  ) |>
  
  fill(
    former_region, .direction = "down"
  ) |> 
  
  # Transform column values
  mutate(
    gender = str_remove_all(gender, "Population, | \\(Number\\)")
  ) |> 
  
  # Remove unwanted rows
  filter(
    !(gender == "Total" | age_category == "Total")
    # gender != "Total", age_category != "Total"
  ) |> 
  
  mutate(
    age_category = case_when(
      str_detect(age_category, "0 to 4|5 to 9|10 to 14") ~ "0-14",
      str_detect(age_category, "15 to 19|20 to 24|25 to 29") ~ "15-29",
      str_detect(age_category, "30 to 34|35 to 39|40 to 44|45 to 49") ~ "30-49",
      str_detect(age_category, "50 to 54|55 to 59|60 to 64") ~ "50-64",
      TRUE ~ "65+"
    )
  ) |> 
  
  # Grouping and summarization
  summarize(
    total = sum(population),
    .by = c(former_region, state, gender, age_category)
  )
```

2.  Writing a Function

    ```{r}
    # Source the function
    box::use(../R/clean)

    file_path <- fs::dir_ls("../00-input/excel_files/")

    # Test the function
    df <- clean$tweak_census(
      path = file_path, 
      pattern = "Population, | \\(Number\\)", 
      condition = !(gender == "Total" | age_category == "Total"),
      group_var = c(former_region, 
                    state, gender, 
                    age_category),
      c(former_region, 
        state = region_name, 
        gender = variable_name, 
        age_category = age_name, 
        population = x2008)
    )
    ```

3.  Grouping and Summarization

```{r}
state <- 
  clean$calc_state_totals(
    df, group_var = c(state, gender)
)
```

2.  Conclusion
